---
title: "Assignment 3"
format: html
editor: visual
---

```{r}
rm(list=ls())
library(tidyverse)
library(dplyr)
library(lubridate)
library(readxl)
```

Task 1

```{r}
library(readxl)
excel_sheets("/Users/vongrov/Documents/sok-1005/ukrainasupport.xlsx")
```

```{r}
library(readxl)
df <- read_excel("ukrainasupport.xlsx", sheet = "Fig 6. With Refugee Support, €", col_names = TRUE)
```

Ploting figure 6

```{r}
colnames(df) <- c("Country", "Total", "Financial", "Humanitarian", "Military", "Refugee_Cost")
df <- df[, !is.na(names(df))]
```

```{r}
df_long$Amount <- as.numeric(df_long$Amount)
```

```{r}
df_long <- df %>%
  pivot_longer(cols = c(Financial, Humanitarian, Military, Refugee_Cost),
               names_to = "Category",
               values_to = "Amount") %>%
  drop_na(Amount, Country)

```

```{r}
ggplot(df_long, aes(x = reorder(Country, Amount), y = Amount, fill = Category)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Financial" = "blue",
                               "Humanitarian" = "red",
                               "Military" = "green",
                               "Refugee_Cost" = "orange")) +
  labs(title = "Total bilateral commitments plus refugee costs (billion Euros)",
       x = "Country", y = "Amount (Billion Euros)",
       fill = "Category") +
  theme_minimal() +
  coord_flip() 
 

```

Task 2

```{r}
url <- "https://raw.githubusercontent.com/ywchiu/rcookbook/master/chapter7/superstore_sales.csv"
data <- read_csv(url)
```

2.1 Showing month with most sales

```{r}
data %>% 
  group_by(`Order Date`) %>% 
  summarise(total_sales = sum(Sales, na.rm = TRUE)) %>% 
  arrange(desc(total_sales)) %>% 
  slice((1))

```

2.2

Showing the province with highest sales

```{r}
data %>% 
  group_by(Province) %>% 
  summarise(total_sales = sum(Sales, na.rm = TRUE)) %>% 
  arrange(desc(total_sales)) %>% 
  slice((1))
```

2.3

```{r}
top_province <- data %>%
  group_by(Province, `Product Category`) %>%  
  summarise(Total_Quantity = sum(`Order Quantity`, na.rm = TRUE)) %>%  
  slice_max(Total_Quantity, n = 1)  

```

Plotter figuren

```{r}
ggplot(top_province, aes(x = Province, y = Total_Quantity, fill = `Product Category`)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Most Sold Product Category per Province",
       x = "Province",
       y = "Total Order Quantity",
       fill = "Product Category") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Task 3

```{r}
library(gapminder)
```

3.1

Minimum and maximum life expectancy values in Asia

```{r}
gapminder %>%
  filter(continent == "Asia") %>%
  group_by(year) %>%
  summarise(min_lifeExp = min(lifeExp), max_lifeExp = max(lifeExp))


```

3.2

Percentage changes using log differentials

```{r}
gapminder %>%
  filter(continent == "Asia") %>%
  group_by(country) %>%
  arrange(year) %>%
  mutate(log_diff_lifeExp = log(lifeExp) - log(lag(lifeExp))) %>%
  summarise(min_growth = min(log_diff_lifeExp, na.rm = TRUE),
            max_growth = max(log_diff_lifeExp, na.rm = TRUE))
```

comparing minimum and maximum growth

```{r}
summary_growth <- gapminder %>%
  filter(continent == "Asia") %>%
  arrange(year) %>%
  mutate(log_diff_lifeExp = log(lifeExp) - log(lag(lifeExp))) %>%
  summarise(overall_min_growth = min(log_diff_lifeExp, na.rm = TRUE),
            overall_max_growth = max(log_diff_lifeExp, na.rm = TRUE))

print(summary_growth)
```

The maximum life expectancy have grown more then minimum life expectancy

3.3

The average life expectancy and GDP

```{r}
gapminder %>%
  group_by(country) %>%
  summarise(avg_lifeExp = mean(lifeExp), avg_gdpPercap = mean(gdpPercap)) %>%
  arrange(desc(avg_lifeExp)) %>%
  top_n(20, avg_lifeExp) %>%
  ggplot(aes(x = reorder(country, avg_lifeExp), y = avg_lifeExp)) +
  geom_point() +
  coord_flip() +
  labs(title = "Top 20 countries based on GDP, with highest average life expectancy",
       x = "Land", y = "Average life expectancy")


```

3.4

Life expectancy VS GDP

```{r}
gapminder %>%
  group_by(country) %>%
  summarise(avg_lifeExp = mean(lifeExp), avg_gdpPercap = mean(gdpPercap), continent = first(continent)) %>%
  ggplot(aes(x = avg_gdpPercap, y = avg_lifeExp, color = continent)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10() +
  labs(title = "Gjennomsnittlig levealder vs BNP per land (log-skala)",
       x = "Gjennomsnittlig BNP per innbygger (log)", y = "Gjennomsnittlig levealder (log)")


```

3.5

Change in life expectancy every 5-years

```{r}
lifeExp_changes <- gapminder %>%
  group_by(country, continent) %>%
  arrange(year) %>%
  mutate(lifeExp_change = lifeExp - lag(lifeExp, 1)) %>%
  select(country, continent, year, lifeExp_change) %>%
  na.omit() 

```

Finding the country with sharpest drop in life expectancy

```{r}
largest_drop <- lifeExp_changes %>%
  group_by(continent) %>%
  filter(lifeExp_change == min(lifeExp_change)) %>%
  arrange(lifeExp_change)

print(largest_drop)


```
